
\scriptsize

\begin{thebibliography}{10}

\bibitem{Atk91}
  A.O.L.~Atkin.
  \newblock \emph{The number of points on an elliptic curve modulo a prime}.
  \newblock Email on the Number Theory Mailing List, 1991.

\bibitem{BK78}R.P.~Brent and H.T.~Kung, Fast algorithms for
  manipulating formal power series.  \emph{J. ACM} 25(4):581--595, 1978.

\bibitem{BlSeSm99}
  I.~Blake, G.~Seroussi, and N.~Smart.
  \newblock {\em Elliptic curves in cryptography}.
  \newblock Cambridge University Press, 1999.

\bibitem{BoMoSaSc08}
  A.~Bostan, F.~Morain, B.~Salvy and É.~Schost
  \newblock Fast algorithms for computing isogenies between elliptic curves.
  \newblock \emph{Math. Comp.} 77, 263:1755-1778, 2008.

\bibitem{Magma}
  W.~Bosma, J.~Cannon, C.~Playoust.
  \newblock The Magma algebra system. I. The user language.
  \newblock \emph{J. Symb. Comp.}, 24(3-4):235-265, 1997.
  
\bibitem{gf2x}
  R.~Brent, P.~Gaudry, E.~Thom{\'e}, P.~Zimmermann.
  \newblock Faster multi\-plication in GF$(2)[x]$.
  \newblock In {\em ANTS'08}, 153-166. Springer, 2008.

\bibitem{Cou94}
  J.-M.~Couveignes.
  \newblock \emph{Quelques calculs en théorie des nombres}.
  \newblock PhD thesis. 1994.

\bibitem{Cou96}
  J.-M.~Couveignes. 
  \newblock Computing {$\ell$}-isogenies using the {$p$}-torsion.
  \newblock in {\em ANTS'II}, 59--65. Springer, 1996.

\bibitem{Cou00}
  J.-M.~Couveignes. 
  \newblock Isomorphisms between {A}rtin-{S}chreier towers. 
  \newblock \emph{Math. Comp.} 69(232): 1625--1631, 2000.
  
\bibitem{DFS09}
  L.~De~Feo and É.~Schost
  \newblock Fast Arithmetics in Artin Schreier Towers.
  \newblock \emph{ISSAC '09}, 2009.

\bibitem{Elk91}
  N.D.~Elikes
  \newblock \emph{Explicit Isogenies}.
  Draft, 1991.

\bibitem{EnMo03}
  A.~Enge and F.~Morain,
  \newblock Fast decomposition of polynomials with known Galois group.
  \newblock in {\em AAECC-15}, 254--264. Springer, 2003.

\bibitem{JL06}A.~Joux, R. Lercier.
  \newblock Counting points on elliptic curves in medium characteristic.
  \newblock \emph{Cryptology ePrint Archive} 2006/176, 2006.

\bibitem{vzGG} 
  J.~von~zur~Gathen and J.~Gerhard. 
  \newblock \emph{Modern Computer Algebra}. 
  \newblock Cambridge University Press, 1999.

\bibitem{Gun76}H.~Gunji,
  \newblock The Hasse Invariant and $p$-division Points of an Elliptic Curve.
  \newblock \emph{Archiv der Mathematik} 27(2), Springer, 1976.

\bibitem{Ler96}
  R.~Lercier.
  \newblock Computing isogenies in GF($2^n$).
  \newblock In {\em ANTS-II}, LNCS vol 1122, pages 197--212. Springer, 1996.

\bibitem{Ler97}
  R.~Lercier.
  \newblock Algorithmique des courbes elliptiques dans les corps finis. 
  \newblock Ph.D. Thesis, {\'E}cole polytechnique, 1997.

\bibitem{LeSi09}
  R.~Lercier, T.~Sirvent.
  \newblock On Elkies subgroups of $\ell$-torsion points in curves defined over
  a finite field.
  \newblock To appear in {\em J. Th\'eor. Nombres Bordeaux}.

\bibitem{RoSt06}
  A.~Rostovtsev and A.~Stolbunov.
  \newblock Public-key cryptosystem based on isogenies.
  \newblock Cryptology ePrint Archive, Report 2006/145.
  
\bibitem{Sat00}
  T.~Satoh.
  \newblock The canonical lift of an ordinary elliptic curve over a nite eld and its point counting.
  \newblock \emph{Journal of the Ramanujan Mathematical Society}, 2000.

\bibitem{Sch95}
  R.~Schoof.
  \newblock Counting points on elliptic curves over finite fields.
  \newblock \emph{J. Théorie des Nombres de Bordeaux} 7:219--254, 1995.

\bibitem{NTL}
  V.~Shoup.
  \newblock {NTL}: A library for doing number theory.
  \newblock \url{http://www.shoup.net/ntl/}.

\bibitem{Tes06}
  E.~Teske.
  \newblock An elliptic trapdoor system.
  \newblock {\em Journal of Cryptology}, 19(1):115--133, 2006.

\bibitem{Vel71}J.~Vélu,
  \newblock Isogénies entre courbes elliptiques.
  \newblock \emph{Comptes Rendus de l'Académie des Sciences de Paris} 273,
  Série A, 238--241, 1971.

  
  
\end{thebibliography}
