\documentclass{article}

\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{bbm}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{mathrsfs}
\usepackage{mdwlist}
\usepackage{graphicx}
\usepackage{mathrsfs}


%% Stuff
\renewcommand{\le}{\leqslant}
\renewcommand{\ge}{\geqslant}  % comme François le demande...
\newcommand{\todo}{\dots}  % un marqueur pour un trou
%% Algèbre
\newcommand{\trans}[1]{#1^\top}  % transposé
\newcommand{\dual}[1]{#1^\ast}  % dual
\newcommand{\clot}[1]{\bar{#1}}  % clôture algèbrique
\newcommand{\card}[1]{\# #1}  % cardinalité d'un ensemble
\DeclareMathOperator{\car}{car}  % caractéristique d'un corps
\DeclareMathOperator{\Frac}{Frac}  % corps des fractions
\newcommand{\N}{\mathbb{N}}  % les naturels
\newcommand{\Z}{\mathbb{Z}}  % les entiers
\newcommand{\K}{\mathbb{K}}  % un corps
\newcommand{\LK}{\mathbb{L}}  % encore un corps
\newcommand{\U}{\mathbb{U}}  % encore un corps
\newcommand{\F}{\mathbb{F}}  % un corps fini
\newcommand{\Q}{\mathbb{Q}}  % les rationnels
\newcommand{\R}{\mathbb{R}}  % les réels
\newcommand{\C}{\mathbb{C}}  % les complexes
\newcommand{\isom}{\cong}  % isomorphisme de corps
\newcommand{\frob}{\varphi}  % l'isomorphisme de Frobenius
\DeclareMathOperator{\Gal}{Gal}  % groupe de Galois
\DeclareMathOperator{\Tr}{Tr}  % trace
\DeclareMathOperator{\PTr}{T}  % pseudotrace
\DeclareMathOperator{\Norm}{N} % norme
\DeclareMathOperator{\coeff}{coeff}  % coefficient
\newcommand{\res}{\rho}  % the residue form
\newcommand{\euler}{\phi}  % indicatrice d'Euler
\DeclareMathOperator{\ord}{ord}  % l'ordre d'un élément
\DeclareMathOperator{\rev}{rev}  % le reverse d'un polynôme
\newcommand{\Cyclo}{\Phi}  % le polynome cyclotomique
\DeclareMathOperator{\Fix}{Fix}  % les points fixes
%% Courbes
\newcommand{\Proj}{\mathbb{P}}  % espace projectif
\newcommand{\0}{\mathcal{O}}  % point de base d'une courbe
\newcommand{\ecpoint}[3]{[#1:#2:#3]}  % un point d'une courbe
\newcommand{\isog}[1]{\mathcal{#1}}  % la police des isogénies
\newcommand{\I}{\isog{I}}  % une isogénie I
\newcommand{\frobisog}{\phi}  % l'isogénie de Frobenius
\newcommand{\Hasse}{H}  % l'invariant de Hasse
\newcommand{\divpol}{f}  % polynôme de division
%% Autre
\newcommand{\tildO}{\tilde{O}}  % la notation O~ qui oublie les log
\newcommand{\Mult}{\mathrm{\sf M}}  % fonction de multiplication
\newcommand{\ModComp}{\mathrm{\sf C}}  % fonction de composition modulaire
\newcommand{\alg}[1]{{\sf #1}}  % la police des algorithmes
\newcommand{\algref}[1]{\alg{\ref{#1}}}  % la police des algorithmes
\newcommand{\wrt}{\dashv}  % appartenance forte, a\wrt A signifie que a est représenté comme un élément de A
\newcommand{\ndiv}{\nmid}  % ne divise pas

%%Théorèmes
\newtheorem{definition}{Definition}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[definition]{Lemma}
\newtheorem{corollary}[definition]{Corollary}
\newtheorem{proposition}[definition]{Proposition}
\newtheorem{problem}[definition]{Problem}


\title{On computing isogenies of unknown degree}
\author{Luca De Feo}

\begin{document}

\maketitle

\begin{abstract}
  We present an extension to Couveignes algorithm that, given two
  elliptic curves $E$ and $E'$, permits to compute all isogenies of
  degrees up to a certain bound $N$ in time $\tildO(N^2)$.
\end{abstract}

It is well known that two curves having the same number of points over
a finite field are isogenous, however this doesn't say anything on the
degree of the isogeny connecting them. Given two elliptic curves $E$
and $E'$ defined over $\F_q$ and having the same number of points, we
want to find the smallest degree isogeny between them.

The simplest solution is to take any algorithm computing a fixed
degree isogeny and try all the degrees until an isogeny is found. If
$\ell$ is the degree of the smallest isogeny, this of course adds a
factor $\ell$ to the complexity of any polynomial time algorithm.

Couveignes' algorithm \cite{Cou96}, however, can be easily adapted to
solve this problem at no additional cost. This short paper is not
self-contained: we make references to the variants and improvements to
Couveignes' algorithm given in \cite{D}.

Observe that in Couveignes' algorithm, apart for the choice of $k$,
the computation of $E[p^k]$ and the polynomial interpolation step do
not depend at all on $\ell$. The degree of the isogeny only comes into
play in the last part of the Cauchy interpolation, that is the
rational function reconstruction. We study more in detail this last
step.

\paragraph{Rational Function Reconstruction}
Rational function reconstruction takes as input a degree $n$
polynomial $T$, a polynomial $A$ of degree less than $n$ and a target
degree $m\le n$ and outputs the unique rational function such that
\begin{equation*}
  A \equiv \frac{R}{V} \bmod T
\end{equation*}
and $\deg R < m$, $\deg V \le n-m$. This is done by computing a Bezout
relation $AV + TU = R$ with the expected degrees via an XGCD
algorithm. If a classical XGCD algorithm is used, one simply computes
all the lines
\begin{equation}
  \label{eq:XGCD}
  \begin{aligned}
    R_0 &= T, & U_0 &= 1, & V_0 &= 0,\\
    R_1 &= A, & U_1 &= 0, & V_1 &= 1,\\
    R_{i-1} &= Q_iR_i + R_{i+1}, & U_{i+1} &= U_{i-1}-Q_iU_i, & V_{i+1} &= V_{i-1}-Q_iV_i
  \end{aligned}
\end{equation}
and stops as soon as a remainder $R_{i+1}$ with $\deg R_{i+1}<m$ is
found. If a fast XGCD algorithm as \cite[Algo. 11.4]{vzGG} is used,
one directly aims at the two lines
\begin{equation}
  \label{eq:FastGCD}
  \begin{aligned}
    R_{h-2} &= Q_{h-1}R_{j-1} + R_h\\
    R_{h-1} &= Q_hR_h + R_{h+1}
  \end{aligned}
\end{equation}
such that $\deg R_{h+1} < m \le \deg R_h$ without computing the
intermediate lines.

When looking for an $\ell$-isogeny, one simply sets
$m=\ell+1$. Observe that if the algorithm doesn't return a rational
fraction $\frac{R}{V}$ with $\deg R = \ell$ and $\deg V = \ell -1 $,
then no such fraction congruent to $A$ modulo $T$ exists.

If $\ell$ is not \emph{a priori} known, we can still use the fact that
a separable isogeny with cyclic kernel must have $\deg R = \deg V +
1$. In fact if we suppose $R = R_i$ and $V = V_i$, then
\begin{align*}
  \deg T &= \deg V_{i+1} + \deg R_i,\\
  \deg R_i - \deg V_i &= \deg R_{i-1} - \deg V_{i+1}
\end{align*}
implies
\begin{equation*}
  \deg T + 1 = \deg R_{i-1} + \deg R_i  
  \;\text{.}
\end{equation*}
Hence, if $A$ is congruent to an $\ell$-isogeny with $\ell =
\left\lfloor\frac{\deg T}{2}\right\rfloor - t$ for some $t\ge0$, then
\begin{equation}
  \label{eq:degseq}
  \deg R_{i-1} =
  \left\lceil\frac{\deg T}{2}\right\rceil + t + 1 >
  \left\lfloor\frac{\deg T}{2}\right\rfloor - t = \deg R_i
  \;\text{.}
\end{equation}
Thus we can recover any isogeny having degree less than
$\left\lfloor\frac{\deg T}{2}\right\rfloor$ using either a classical
or a fast XGCD algorithm setting $m = \left\lceil\frac{\deg
    T}{2}\right\rceil + 1$.


\paragraph{Recognising an isogeny}
Once we have a rational fraction with the required degree, we have to
test if it really is an isogeny. In order to understand how often we
have to make this test, we introduce some more terminology. Let $n_i =
\deg R_i$, we call $(n_0,\ldots,n_r)$ the \emph{degree sequence} of
$A$ and $T$; a degree sequence is said \emph{normal} if $n_i = n_{i+1}
+ 1$ for any $i$.

\begin{proposition}
  \label{th:normseq}
  Let $f,g\in\F_q[X]$ be uniformly chosen random polynomials of
  respective degrees $n_0>n_1>0$ and let $(n_0, n_1, \ldots, n_r)$ be
  their degree sequence. For $0\le i < n_1$ define the binary random
  variables $X_i = 1 \Leftrightarrow i\in(n_0,n_1,\ldots,n_r)$, then
  the $X_i$ are independent random variables and $\mathrm{Prob}(X_i=0) =
  \frac{1}{q}$.
\end{proposition}
\begin{proof}
  Pairs of polynomials $f,g$ are in bijection with the GCD-sequence
  $(R_r, Q_r, \ldots, Q_1)$ constituted by their GCD and the quotients
  of the GCD algorithm. To each such sequence is associated a degree
  sequence
  \begin{equation*}
    (n_0,n_1,\ldots,n_r) =
    \left(\deg R_r + \sum_{i=1}^r\deg Q_i, \ldots, \deg R_r + \sum_{i=1}^1\deg Q_i, \deg R_r\right)
    \;\text{,} 
  \end{equation*}
  thus for any given degree sequence there are
  \begin{equation*}
    (q-1)q^{n_0-n_1}\cdot(q-1)q^{n_1-n_2}\cdot\cdots\cdot(q-1)q^{n_r} =
    (q-1)^{r+1}q^{n_0}
  \end{equation*}
  GCD-sequences.
 
  Let $I$ and $O$ be two disjoints subsets of $\{X_i\}$, the number of
  GCD-sequences such that $X\in I \Rightarrow X=1$ and $X\in O
  \Rightarrow X=0$,
  \begin{equation*}
     \sum_{s=0}^{n_1-\card{I}-\card{O}}\binom{n_1-\card{I}-\card{O}}{s}(q-1)^{s+2+\card{I}}q^{n_0} =
    (q-1)^{2+\card{I}}q^{n_0}q^{n_1-\card{I}-\card{O}}
    \;\text{.}
  \end{equation*}
  There are $(q-1)^2q^{n_0}q^{n_1}$ pairs of polynomials of degrees
  $n_0,n_1$, thus
  \begin{equation}
   \label{th:normseq:prob}
    \mathrm{Prob}\bigl(\{X = 1 \mid X\in I\},
    \{X=0\mid X\in O\}\bigr) = \left(\frac{q-1}{q}\right)^{\card{I}}\left(\frac{1}{q}\right)^{\card{O}}
    \;\text{.}
  \end{equation}
  The claim follows.
\end{proof}

Degree sequences associated to isogenies are in general not normal, in
fact if $\ell\le\left\lfloor\frac{\deg T}{2}\right\rfloor-t$, equation
\eqref{eq:degseq} shows that there must be at least a gap of degree
$2c$ in the degree sequence. Heuristically, we can expect that if the
polynomial $A$ doesn't correspond to an isogeny, then $A$ and $T$ act
like random polynomials, thus, by the proposition above, the
probability that $A$ looks like an isogeny of degree
$\ell\le\left\lfloor\frac{\deg T}{2}\right\rfloor-t$ is less than $\frac{1}{q^{2t}}$.

Therefore, by choosing an appropriate $t\in O(\log_q p^k)$, our
variant can find any isogeny of degree less than $\frac{p^k-1}{4}-t$
at the same cost of one run of Couveignes' algorithm.

No other method for computing isogenies is known to have a similar
generalisation. This makes Couveignes' algorithm a rather surprising
exception and we wonder whether this simple idea can find interesting
applications.

\begin{thebibliography}{9}
\bibitem{Cou96}
  J.-M.~Couveignes. 
  \newblock Computing {$\ell$}-isogenies using the {$p$}-torsion.
  \newblock in {\em ANTS'II}, 59--65. Springer, 1996.

\bibitem{D}
  L.~De~Feo.
  \newblock Fast algorithms for computing isogenies between ordinary
  elliptic curves in small characteristic,
  \newblock To appear in {\em Journal of Number Theory}, 2010.

\bibitem{vzGG} 
  J.~von~zur~Gathen and J.~Gerhard. 
  \newblock \emph{Modern Computer Algebra}. 
  \newblock Cambridge University Press, 1999.

\end{thebibliography}


\end{document}

% Local Variables:
% mode:flyspell
% ispell-local-dictionary:"british"
% mode:TeX-PDF
% End:
%
% LocalWords:  Schreier Artin pseudotrace Frobenius bivariate Joux Sirvent FFT
% LocalWords:  Couveignes isogenies Schoof isogeny cryptosystems Lercier
% LocalWords:  precomputation arithmetics polylogarithmic Karatsuba precomputes
% LocalWords:  endomorphisms  isogenous
